---
title: "sparrow_project"
date: today

toc: true
abstract-title: "Research question:"
abstract: "How do insect abundance and ambient noise levels affect the number of fledglings per nest in house sparrows?"

format:
  html:
    embed-resources: true
    number-sections: true
    code-fold: show

editor: visual
editor_options:
  chunk_output_type: console

execute:
  warning: false
  message: false
---

## Prepare Session

```{r}
#| label: prepare-session
#| code-fold: false


#importing libraries
  library(pastecs)
  library(ggplot2)
  library(ggpubr); theme_set(theme_bw())
  library(GGally)
  library(performance)
  library(MuMIn)
  library(ggeffects)
  library(tidyverse)

  library(knitr)  

#import data

library(here)

clean <- read_delim(
  here("data", "processed", "clean.csv"),
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)


```

## Explore Data

### Numerically

```{r}
#| label: tbl-explore-data
#| tbl-cap: "Descriptive statistics for the variables of interest."
#| code-fold: true

sum_stats<-
   stat.desc(clean[, c("num_fledglings", "insect_abundance", "noise")], norm= T)

kable(sum_stats, digits=1)
```

### Graphically

```{r}
#| label: fig-scatterplot-matrix
#| fig-cap: "Scatterplopt matrix of the sparrow data"
#| output: true

# Scatterplot matrix
  ggpairs(clean[, c("num_fledglings", "insect_abundance", "noise")])
```

```{r}
#| label: fig-raw-data
#| fig-cap: "The number of fledglings per nest plotted against the two independent variables (a) insect abundance, and b) environmental noise)."
#| fig-width: !expr 16/2.54
#| fig-height: !expr 9/2.54
#| code-fold: true

# Draw a graph that captures the research question
  ggarrange(
    ggplot(clean, aes(insect_abundance, num_fledglings)) +
      geom_smooth(method= glm, method.args= list(family= "poisson"),
                  col= "black") +
      labs(x= expression("insect abundance (g/m^2)"),
           y= "number of fledglings\n per nest "),
    ggplot(clean, aes(noise, num_fledglings)) +
      geom_smooth(method= glm, method.args= list(family= "poisson"),
                  col= "black") +
      labs(x= "Environmental noise (dB)",
           y= "Number of fledglings\n per nest"),
    nrow= 1, ncol= 2, common.legend= T, labels= c("a)", "b)"))
```

## Fit models

### Formulate and compare

As seen in @tbl-fit-models further analysis with `glm.01` seems like the safest choice. This choice is based on the lowest AIC score and the highest Weight score.

```{r}
#| label: tbl-fit-models
#| tbl-cap: "Model comparison."

# Fit binomial glms
  glm.00<- glm(num_fledglings ~ 1, family= "poisson", clean)
  glm.01<- glm(num_fledglings ~ insect_abundance + noise,
               family= "poisson", clean)
  glm.02<- glm(num_fledglings ~ insect_abundance * noise, 
               family= "poisson", clean)

# Compare models
  compare_performance(glm.00, glm.01, glm.02)
  kable(compare_performance(glm.00, glm.01, glm.02), digits= 3)

```

### Validate preferred model

```{r}
#| label: diagnostics-overdispersion

# First thing: check overdispersion
  check_overdispersion(glm.01)
```

no overdispersion detected

```{r}
#| label: fig-diagnostic-plots
#| fig-cap: "Model diagnostics"
#| fig-width: !expr 25/2.54
#| fig-height: !expr 30/2.54

# Validate model
  check_model(glm.01)
```

The diagnostic plot in @fig-diagnostic-plots does not look too bad.

## Interpret preferred model

```{r}
#| label: interpret-objects
#| output: false

# Objects to construct table ------------------------------------------------

## Coefficients, SE, z and p -----
  coef.table<- round(summary(glm.01)$coefficients, 3)

## Approximate R2 -----
  r.2<- round(attributes(r.squaredLR(glm.01))$adj.r.squared, 3)  

## Compare performance to null-model -----
  null.chi<- anova(glm.00, glm.01)

# Objects to plot predictions -----------------------------------------------

## Interaction plot -----
  
  # Create predictions dataframe
  pred.int<- predict_response(glm.01,
                              terms =c("insect_abundance","noise"))
  
  
  # Create plot
  pred.plot<- ggplot(pred.int, aes(x, predicted)) +
    geom_ribbon(aes(ymin= conf.low, ymax= conf.high,
                    fill= group), alpha= .2) +
    geom_line(aes(col= group)) +
    labs(x= "Insect abundance (g/m^2)",
         y= "Number of fledglings\n per nest", col= "noise", fill= "noise") +
    coord_cartesian(ylim = c(0, max(pred.int$conf.high)))
pred.plot
```

```{r}
#| label: tbl-glm
#| tbl-cap: !expr paste("Estimates, associated standard errors and statistical significance of the parameters estimated by a poisson GLM.", "($R^2$=", r.2, ", $\\chi^2_{(", null.chi[2, 3], ",", null.chi[2, 1], ")}$=", round(null.chi[2,4], 3), ", $p$", format.pval(null.chi[2, 5], eps = 0.001), ")")

# Create table
  kable(coef.table)
```
