---
title: "sparrow_project"
date: today

toc: true
abstract-title: "Research question:"
abstract: "How do insect abundance and ambient noise levels affect the number of fledglings per nest in house sparrows?"

format:
  html:
    embed-resources: true
    number-sections: true
    code-fold: show

editor: visual
editor_options:
  chunk_output_type: console

execute:
  warning: false
  message: false
---

## Prepare Session

For this exercise all the code will be available either directly or folded.

```{r}
#| label: prepare-session
#| code-fold: true


#importing libraries
  library(pastecs)
  library(ggplot2)
  library(ggpubr); theme_set(theme_bw())
  library(GGally)
  library(performance)
  library(MuMIn)
  library(ggeffects)
  library(tidyverse)

  library(knitr)  

#import data

library(here)

clean <- read_delim(
  here("data", "processed", "clean.csv"),
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)


```

## Explore Data

### Numerically

```{r}
#| label: tbl-explore-data
#| tbl-cap: "Descriptive statistics for the variables of interest."
#| code-fold: true

sum_stats<-
   stat.desc(clean[, c("num_fledglings", "insect_abundance", "noise")], norm= T)

kable(sum_stats, digits=1)
```

### Graphically

```{r}
#| label: fig-scatterplot-matrix
#| fig-cap: "Scatterplopt matrix of the sparrow data"
#| output: false
#| code-fold: true

# Scatterplot matrix
  ggpairs(clean[, c("num_fledglings", "insect_abundance", "noise")])
```

Looking at @fig-raw-data following assumptions can be made:

a\) Insect abundance (left panel)

There is a clear positive relationship between insect abundance and number of fledglings. This leads to the interpretation that higher food availability is associated with increased reproductive success.

The confidence band broadens at higher values meaning more uncertainty probably because of fewer observations.

b\) Environmental noise (right panel)

There is a clear negative relationship between environmental noise and number of fledglings. This fact leads to the interpretation that higher environmental noise is associated with reduced reproductive success.

```{r}
#| label: fig-raw-data
#| fig-cap: !expr paste("The number of fledglings per nest plotted against the two independent variables (a) insect abundance (g/$m^2$), and b) environmental noise).")
#| fig-width: !expr 16/2.54
#| fig-height: !expr 9/2.54
#| code-fold: true

# Draw a graph that captures the research question
  ggarrange(
    ggplot(clean, aes(insect_abundance, num_fledglings)) +
      geom_smooth(method= glm, method.args= list(family= "poisson"),
                  col= "black") +
      labs(x= expression("insect abundance (g/" * m^2 * ")"),
           y= "number of fledglings\n per nest "),
    ggplot(clean, aes(noise, num_fledglings)) +
      geom_smooth(method= glm, method.args= list(family= "poisson"),
                  col= "black") +
      labs(x= "Environmental noise (dB)",
           y= "Number of fledglings\n per nest"),
    nrow= 1, ncol= 2, common.legend= T, labels= c("a)", "b)"))
```

The non-linear nature of both relationships (curvature) supports the use of a generalized linear model rather than simple linear regression.

## Fit models

### Formulate and compare

As seen in @tbl-fit-models further analysis with `glm.01` seems like the safest choice. This choice is based on the lowest AIC score and the highest Weight score.

```{r}
#| label: tbl-fit-models
#| tbl-cap: "Model comparison."

# Fit binomial glms
  glm.00<- glm(num_fledglings ~ 1, family= "poisson", clean)
  glm.01<- glm(num_fledglings ~ insect_abundance + noise,
               family= "poisson", clean)
  glm.02<- glm(num_fledglings ~ insect_abundance * noise, 
               family= "poisson", clean)

# Compare models
  compare_performance(glm.00, glm.01, glm.02)
  kable(compare_performance(glm.00, glm.01, glm.02), digits= 3)

```

### Validate preferred model

```{r}
#| label: diagnostics-overdispersion
#| code-fold: true

# First thing: check overdispersion
  check_overdispersion(glm.01)
```

No overdispersion in the data detected.

```{r}
#| label: fig-diagnostic-plots
#| fig-cap: "Model diagnostics"
#| fig-width: !expr 25/2.54
#| fig-height: !expr 30/2.54

# Validate model
  check_model(glm.01)


```

The diagnostic plot in @fig-diagnostic-plots indicates a mismatch between observed and expected variance, suggesting underdispersion in the data. For validation purposes another model is being checked. Instead of a Poisson model negative binomial (\@fig-diagnostic-plots-2) model is tested.

```{r}
#| label: fig-diagnostic-plots-2
#| fig-cap: "Model diagnostics"
#| fig-width: !expr 25/2.54
#| fig-height: !expr 30/2.54

library(MASS)

glm.nb.01 <- glm.nb(num_fledglings ~ insect_abundance + noise,
                    data = clean)

compare_performance(glm.01, glm.nb.01)

check_model(glm.nb.01)


```

A negative binomial model did not improve model fit. Given that the Poisson model remained the best-performing model based on AIC and the alternative model did not provide improvement, the Poisson GLM was kept for further analysis.

## Interpret preferred model

```{r}
#| label: interpret-objects
#| output: false
#| code-fold: true

# Objects to construct table ------------------------------------------------

## Coefficients, SE, z and p -----
  coef.table<- round(summary(glm.01)$coefficients, 3)

## Approximate R2 -----
  r.2<- round(attributes(r.squaredLR(glm.01))$adj.r.squared, 3)  

## Compare performance to null-model -----
  null.chi<- anova(glm.00, glm.01)

# Objects to plot predictions -----------------------------------------------

## Interaction plot -----
  
  # Create predictions dataframe
  pred.int<- predict_response(glm.01,
                              terms =c("insect_abundance","noise"))
  
  
  # Create plot
  pred.plot<- ggplot(pred.int, aes(x, predicted)) +
    geom_ribbon(aes(ymin= conf.low, ymax= conf.high,
                    fill= group), alpha= .2) +
    geom_line(aes(col= group)) +
    labs(x= expression("Insect abundance (g/" * m^2 *")"),
         y= "Number of fledglings\n per nest", col= "noise", fill= "noise") +
    coord_cartesian(ylim = c(0, max(pred.int$conf.high)))
```

```{r}
#| label: tbl-glm
#| tbl-cap: !expr paste("Estimates, associated standard errors and statistical significance of the parameters estimated by a poisson GLM.", "($R^2$=", r.2, ", $\\chi^2_{(", null.chi[2, 3], ",", null.chi[2, 1], ")}$=", round(null.chi[2,4], 3), ", $p$", format.pval(null.chi[2, 5], eps = 0.001), ")")

# Create table
  kable(coef.table)
```

```{r}
#| label: fig-predict
#| fig-cap: "Raw data and model predictions to visualize the relationship between the number of sparrow fledglings per nest, and the insect abundance and environmental noise."
#| fig-width: !expr 15/2.54
#| fig-height: !expr 9/2.54

# Plot predictions
  pred.plot
```
